Spatial and temporal variability in economic availability of residual biomass

### 2021-11-23 analysis/manuscript.Rmd
after els

```{r libraries, echo=FALSE, message=FALSE}
#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra) 
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
library(quantmod)
library(bookdown)
```

@@

1. **Focus on cost variability associated with transportation and forest  characteristics**
2. **Based on historical harvest locations (1990-2015)**

Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020a.  Our
strengths

@@

## Justification

@@

Biofuels are becoming an increasingly important source of energy because
of their potential to reduce dependency on fossil fuels, thereby
reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
generation biofuel provides the benefit of avoiding competition with
food production by using lignocellulosic biomass -- often in the form of
agriculture and forest residues -- as feedstock (source?). Forest
residues are generated as by-products of conventional forest harvesting
operations, and include the tree tops, branches and non-merchantable
logs. These residues are typically left behind in slash piles, which are
burned during winter to mitigate against the risk of wildfires. Canada
has a developed forestry sector that produces approximately 20 million
oven-dry metric tons  of forest residues (Dymond et al. 2010) @Dymond2010.

Despite the potential to produce biofuels from forest residues, such use
is limited to a few highly subsidized pilot projects. Indeed, several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and markets,
and (3) concerns surrounding feedstock availability (Shabani et al.
2013) @Shabani2013. The purpose of our study is to explore concerns surrounding this
third challenge -- feedstock availability -- by evaluating annual
variability in commercial stemwood harvest in the province of Alberta,
Canada.

A number of studies have shown that collection and transportation
costs are arguably the biggest constraint to widespread use of forest
residues in bioenergy systems (Ralevic et al 2003, Wood and Layzell
2003, Kumar et al. 2003, Rummer 2007, Aulakh 2008)
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a biorefinery largely depend on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products could have a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities (de Avillez 2014, Spelter 2009)
@deAvillez2014 @Spelter2009. This relationship between volatile sawlog
harvests and subsequent residue variability is explored in detail by
Niquidet and Friesen (2014) @Niquidet2014, who model forest residue
supply in Alberta as a direct function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a biorefinery's total
costs (Stephen et al. 2010) @Stephen2010, reliable feedstock supplies
and costs over the life of the biorefinery are important factors to
consider when deciding where to locate the biorefinery.

Though past studies regarding biorefinery locations have greatly
improved our understanding of feedstock availability and constraints,
one key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based biorefinery is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a biorefinery

Although not related to a biorefinery siting decision, some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
Moiseyev et al. 2011) @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries (Cambero et al 2015, Chen and Fay 2011, Papapostolou and
Kondili 2011) @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
biorefinery feedstocks (Papapostolou and Kondili 2011)
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities (Cambero et al.
2015) @Cambero2015.

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a biorefinery.

We consider co-locating the biorefinery adjacent to either a sawmill or
a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many of
the costs associated with establishing a new biorefinery (Browne et al.,
2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
2007). @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to co-locate a biorefinery near a sawmill or
pulpmill should be informed by how stable the respective residue
supplies are. The facility with a more stable stemwood harvest would
generate a more stable supply of forest residues that could be collected
from the nearby harvested areas.

@@

The remainder of the paper is as follows: The next section describes
.......

**OTHER POTENTIALLY USEFUL TIDBITS..**

The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low (Golecha and Gan 2016) @Golecha2016.

Modern biorefineries can expect to breakeven financially with delivered
biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
(Steeper 2021) @Steeper2021 . Studies have shown that forest policy reforms could
reduce delivered forest residue costs by incenting plantations of
fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012 @Anderson2012). Other important
factors will be advances in biorefinery technology and increases in
biofuel prices. In Canada, governments impact biofuel prices by setting
renewable fuel mandates and paying subsidies for biofuel production
(Campbell et al., 2016). The continuation of these mandates and
subsidies will likely be an important factor for the emergence of a
second-generation ("advanced") biofuel sector. Also, the Government of
Canada\'s Clean Fuel Standard, which is proposed to come into force in
2022 (Government of Canada, 2020), is expected to improve the economics
of advanced biofuel production.
`

>>>>

```{r libraries, echo=FALSE, message=FALSE}
#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

@@

@@

@@

Humans have been using biofuels since we learned to control fire a
million or so years ago.  In developed economies, biofuels were
largely displaced by coal during the industrial revolution, and by
petroleum starting in the mid-nineteenth century.  In recent times,
biofuels are an increasingly important source of energy because of
their potential to reduce dependency on fossil fuels and reduce
greenhouse gas emissions.  Direct combustion of biomass is still an
important source of energy.  In Europe, in particular, many
electricity plants have been converted to use wood pellets instead of
coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
and biodiesel) have been produced in commercial quantities from food
crops starting in the 1980s.  These are called first generation
biofuels.  A major difficulty with these first generation biofuels is
that they compete with food and that they may lead to land use change.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels.  An advantage of this process is that the feedstock can be
residue from food or forest products production.  Second generation
biofuel production is still largely in the pilot phase.  An
interesting example using forest residues is the Silva Green Fuel
renewable crude oil plant in Tofte, Norway which is expected to begin
operations in 2021 @Lane2018.

Wood pellets are a well-established biofuel.  It is projected that
Canada will produce 3.8 million Mg of wood pellets in 2021, of which
87% will be exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill waste.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

<!-- Clean fuel standard https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html  -->

<!-- Despite the potential to produce biofuels from forest residues, such use -->
<!-- is limited to a few highly subsidized pilot projects. Several -->
<!-- operational and economic challenges hinder the utilization of forest -->
<!-- residues. These challenges are related to: (1) capital investments in -->
<!-- biorefineries, (2) geographical distance between feedstock and markets, -->
<!-- and (3) concerns surrounding feedstock availability @Shabani2013. The purpose of our study is to explore concerns surrounding this -->
<!-- third challenge -- feedstock availability -- by evaluating annual -->
<!-- variability in commercial stemwood harvest in the province of Alberta, -->
<!-- Canada. -->

A number of studies have shown that collection and transportation
costs are arguably the biggest constraint to widespread use of forest
residues in bioenergy systems
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depend on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs @Stephen2010, reliable feedstock supplies
and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Though past studies regarding bioenergy plant locations have greatly
improved our understanding of feedstock availability and constraints,
one key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant

Although not related to a bioenergy plant siting decision, some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
 @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
bioenergy plant feedstocks (Papapostolou and Kondili 2011)
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities  @Cambero2015.

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a bioenergy plant.

We consider co-locating the bioenergy adjacent to either a sawmill
or a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many
of the costs associated with establishing a new bioenergy plant
@Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to
co-locate a bioenergy plant near a sawmill or pulpmill should be informed
by how stable the respective residue supplies are. The facility with a
more stable stemwood harvest would generate a more stable supply of
forest residues that could be collected from the nearby harvested
areas.

@@

The remainder of the paper is as follows: The next section describes

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimations for timber
supply areas in British Columbia (see Ref. @Blackburn2020 for an
example).  Like us, FPInnovations develops supply curves
(cost-availability curves based on a road network and historical
harvests).  However, their take on temporal variability is limited to
examining 2 5-year periods. As well, they limit their analyses to the
boundaries of timber supply areas.  We do not limit ourselves with
arbitrary boundaries.  Supply cost curves.

Our analysis also shares some similarities with work published by
Yemshanov *et al.* @Yemshanov2014. National level based on costs to
cogeneration facilities. Potential harvest locations.   Again our interest included
temporal variability. actual harvest locations.
<!-- ....... -->

<!-- **OTHER POTENTIALLY USEFUL TIDBITS..** -->

<!-- The need to go outside its normal delivery zone in bad years is similar -->
<!-- to a scenario in the agriculture residue feedstock supply literature, -->
<!-- called the the "derisked" scenario, whereby a buffer area outside the -->
<!-- normal supply zone is contracted to provide residues during years when -->
<!-- residue supply is low @Golecha2016. -->

<!-- Modern biorefineries can expect to breakeven financially with -->
<!-- delivered biomass costs of approximately \$80 (US\$60) t\[oven -->
<!-- dried\]^-1^ (Steeper 2021) @Steeper2021 . Studies have shown that -->
<!-- forest policy reforms could reduce delivered forest residue costs by -->
<!-- incenting plantations of fast-growing tree species on lands close to -->
<!-- pulpmill locations (Shooshtarian et al. 2021 @Shooshtarian2021, -->
<!-- Anderson et al. 2012 @Anderson2012). Other important factors will be -->
<!-- advances in biorefinery technology and increases in biofuel prices. In -->
<!-- Canada, governments impact biofuel prices by setting renewable fuel -->
<!-- mandates and paying subsidies for biofuel production (Campbell et al., -->
<!-- 2016). The continuation of these mandates and subsidies will likely be -->
<!-- an important factor for the emergence of a second-generation -->
<!-- ("advanced") biofuel sector. Also, the Government of Canada\'s Clean -->
<!-- Fuel Standard, which is proposed to come into force in 2022 -->
<!-- (Government of Canada, 2020), is expected to improve the economics of -->
<!-- advanced biofuel production.   -->`
